Relationships in the extremes and their influence on competition and coexistence

### 2022-04-28 paper/Paper.Rnw
minor tweaks

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which actually matter for coexistence.
Vital rates are often measured directly in demographic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not yet been done.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms result from rendering one or more of $X$, $Y$, and $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present difficulties.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}. We have here used simple mathematical
and laboratory models to provide an initial exploration of the potential importance of ATAs
for coexistence. Future work should attempt to generalize lessons learned here to 
field systems with multiple species, stage structure, spatial extent, non-competitive interactions 
such as predation and facilitation that can also mediate coexistence, and other complexities.

@@

The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, and Robin Snyder for helpful discussions. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, and the California Department of Fish and
Wildlife Delta Science Program. 
%***DAN: DAN: if we take Ellner off the author list add him back here, and also add him back to the acks in the README.

>>>>

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', 
in fact it is often instead a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is probably better conceptualized as a species response.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
So the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which matter for coexistence.
Vital rates are measured directly in demographic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not been done.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.\label{eq:11}
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms in (\ref{eq:11}) result from rendering one or 
more of $X$, $Y$, and $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present challenges.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}. We have here used simple mathematical
and laboratory models to provide an initial exploration of the potential importance of ATAs
for coexistence. Future work should attempt to generalize lessons learned here to 
field systems with multiple species, stage structure, spatial extent, non-competitive interactions 
such as predation and facilitation that can also mediate coexistence, and other complexities.

@@

The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, and Robin Snyder for helpful discussions. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, and the California Department of Fish and
Wildlife Delta Science Program. 
%***DAN: DAN: if we take Ellner off the author list add him back here, and also add him back to the acks in the README. Also take him off the author list in sup mat.
